Quantitative Evaluation of Touch-Screen Performance with Transition Matrix Measures

### 2015-09-01 flux-entropy-paper/flux-entropy-paper.tex
-
fixed up a bit of the flux def

It has been recognised that collaborative interactions such as
improvised musical performances can be modelled as a sequence of
states. In the following section we present a method for analysing the
transitions between adjacent states in such a sequence and derive two
high-level quantities that we use to compare musical performances
under different circumstances. Suppose that an collaborative
interaction can be characterised by a set of interaction states $S$
with $m$ members. Each participant's activity can then be represented
as a sequence:
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
Where each $X_i$ is a member of $G$. The transitions between different
interaction states can be summarised in a transition matrix as if the
sequence $X_n$ were a first-order Markov chain. Let $N_{ij}$ be the
number of times that state $j$ follows state $i$. Each participant's
transition matrix, $P$, will then be an $m \times m$ matrix with
entries calculated thus
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the transition matrix for each performer.

To compare the transition matrices of different interactions we use
two high-level quantities, ``flux'' and ``entropy''. The flux of a
sequence is defined as the ratio of changed transitions (e.g. state $x$
to state $y$), to self transitions (e.g. state $x$ to state $x$). The
the flux of a transition matrix $P$ can be calculated as:
\begin{equation}
  \mathrm{flux}(P) = 1 - tr(P)
\end{equation}
Where the trace of P, $tr(P)$ is given by the vector 1-norm of the
diagonal entries of $P$. 
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently a group changes state. Flux
returns a value in the interval $[0,1]$ and will return 0 when
participants never change state, and 1 when participants never stay on
the same state for two measurements in a row.

A related (but different) high-level derived quantity for the
transition matrix is its entropy, defined in the
information-theoretic~\cite{Shannon:1948rt} sense:
\begin{equation}
  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
\end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction,
on the other hand, will be low. Entropy, therefore, is a measure of
how broad the participant's exploration of the state space is in a given
interaction.

@@

% todo need to add overall contrasts column to do pairwise t-tests.

>>>>

It has been recognised that collaborative interactions such as
improvised musical performances can be modelled as a sequence of
states. In the following section we present a method for analysing the
transitions between adjacent states in such a sequence and derive two
high-level quantities that we use to compare musical performances
under different circumstances. Suppose that an collaborative
interaction can be characterised by a set of interaction states $S$
with $m$ members. Each participant's activity can then be represented
as a sequence:
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
Where each $X_i$ is a member of $G$. The transitions between different
interaction states can be summarised in a transition matrix as if the
sequence $X_n$ were a first-order Markov chain. Let $N_{ij}$ be the
number of times that state $j$ follows state $i$. Each participant's
transition matrix, $P$, will then be an $m \times m$ matrix with
entries calculated thus
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the transition matrix for each performer.
The matrix is normalised so that the sum of all elements will be one
(rather than the sum of each row being one, as for a right-stochastic matrix).

To compare the transition matrices of different interactions we use
two high-level quantities, ``flux'' and ``entropy''. The flux of a
sequence is defined as the ratio of changed transitions (e.g.
$A \rightarrow B$), to self transitions (e.g. $A \rightarrow A $). The
the flux of a transition matrix $P$ can easily be calculated:
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
Where the trace of P, $tr(P)$ is given by the sum of the diagonal
entries of $P$.
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently a group changes state. Flux
returns a value in the interval $[0,1]$ and will return 0 when
participants never change state, and 1 when participants never stay on
the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic~\cite{Shannon:1948rt}
sense:
\begin{equation}
  H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})
\end{equation}
This entropy measure is small when the matrix is sparse, and largest
when each matrix element is equal. It offers a different perspective
on collaborative interactions than the flux measure by capturing the
breadth of the gestural space explored by the participants throughout
the course of an interaction. Consider the degenerate case of a
participant who alternates between two states for a whole performance:
the flux in this case will be maximal ($\mathrm{flux}(P) = 1$) since
there are no self-transitions, (only $A \rightarrow B$ or
$ B \rightarrow A$) even though the participant has only used a small
subset of the state space. The entropy of this interaction,
on the other hand, will be low. Entropy, therefore, is a measure of
how broad the participant's exploration of the state space is in a given
interaction.

@@

\begin{itemize}
\item todo: need to add overall contrasts column to do pairwise t-tests.
\end{itemize}
